---
title: "NFL Salary Modeling by Position Group"
author: "Jacob Cohen and Michael Nichols"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# 1. Load Packages and Data

```{r}
library(tidyverse)
library(knitr)    # tables
library(broom)    # tidy model output
library(ggplot2)  # plots

# File paths (assumes the .Rmd lives in the same folder as the CSVs)
Offense <- read.csv("merged_players_offense_yearly_combined.csv")
Defense <- read.csv("merged_players_defense_yearly_combined.csv")
```

# 2. Data Filtering and Position Groups

In this section we restrict the data to the most recent 12 seasons and create
broad position groups that roughly align with how front offices value players:
quarterbacks (QB), running backs (RB), pass-catchers (WR/TE), and defense.

```{r}
# Determine last season year in each dataset
last_year_off <- max(Offense$season__yr_off, na.rm = TRUE)
last_year_def <- max(Defense$season__yr_def, na.rm = TRUE)

cutoff_off <- last_year_off - 11
cutoff_def <- last_year_def - 11

# Offense: keep positive contract values and recent seasons
offense_clean <- Offense %>%
  filter(
    !is.na(value__cmb),
    value__cmb > 0,
    season__yr_off >= cutoff_off,
    season__yr_off <= last_year_off
  ) %>%
  mutate(
    # approximate age from birth year
    age = season__yr_off - birth_year__yr_off,
    age = if_else(age < 18 | age > 45, NA_real_, age)
  ) %>%
  drop_na(age)

# Production‑based position groups
offense_pos <- offense_clean %>%
  mutate(
    pos_group = case_when(
      # QBs: have non‑trivial passing yards
      season_passing_yards__yr_off > 0 ~ "QB",
      # RBs: rushing‑focused, little receiving volume
      season_rushing_yards__yr_off > 0 &
        season_receiving_yards__yr_off < 300 ~ "RB",
      # WR + TE: primary pass catchers
      season_receiving_yards__yr_off >= 300 ~ "WR_TE",
      TRUE ~ "Other"
    )
  ) %>%
  filter(pos_group != "Other")

table(offense_pos$pos_group)
```

For defense we keep the same 12‑year window and create basic performance
variables: tackles, sacks, hits, and interceptions.

```{r}
defense_clean <- Defense %>%
  filter(
    !is.na(value__cmb),
    value__cmb > 0,
    season__yr_def >= cutoff_def,
    season__yr_def <= last_year_def
  ) %>%
  mutate(
    age       = age__yr_def,
    years_exp = seasons_played__yr_def,
    log_value = log(value__cmb),
    total_tackles = tackle_with_assist__yr_def,
    sacks         = sack__yr_def,
    qb_hits       = qb_hit__yr_def,
    ints          = interception__yr_def
  ) %>%
  drop_na(age, years_exp, log_value,
          total_tackles, sacks, qb_hits, ints)
```

# 3. Helper Functions

```{r}
# Z‑score standardization
standardize <- function(x) as.numeric(scale(x))

# RMSE and R²
regression_metrics <- function(actual, predicted) {
  rmse <- sqrt(mean((predicted - actual)^2))
  ss_res <- sum((actual - predicted)^2)
  ss_tot <- sum((actual - mean(actual))^2)
  r2   <- 1 - ss_res / ss_tot
  tibble(RMSE = rmse, R2 = r2)
}

# Train / test / validation split
split_three_way <- function(dat, train_prop = 0.6, test_prop = 0.2, seed = 340) {
  stopifnot(train_prop + test_prop < 1)
  set.seed(seed)
  n <- nrow(dat)
  idx_all <- sample(seq_len(n), size = n, replace = FALSE)

  n_train <- floor(train_prop * n)
  n_test  <- floor(test_prop * n)

  train_idx <- idx_all[1:n_train]
  test_idx  <- idx_all[(n_train + 1):(n_train + n_test)]
  val_idx   <- idx_all[(n_train + n_test + 1):n]

  list(
    train = dat[train_idx, , drop = FALSE],
    test  = dat[test_idx,  , drop = FALSE],
    val   = dat[val_idx,   , drop = FALSE]
  )
}
```

# 4. Offensive Regression Models by Position

We fit separate linear regressions within each offensive position group.  
The response is log(contract value), so coefficients can be interpreted as
percentage changes in contract value for a one‑standard‑deviation increase in
a predictor.

```{r}
fit_position_model <- function(dat, pos_label) {
  message("----- Fitting model for position group: ", pos_label)

  dat_pos <- dat %>%
    filter(pos_group == pos_label) %>%
    mutate(
      years_exp = years_exp__yr_off,
      pass_yards = season_passing_yards__yr_off,
      pass_tds   = season_pass_touchdown__yr_off,
      ints       = interception__yr_off,
      pass_att   = season_pass_attempts__yr_off,
      rush_yards = season_rushing_yards__yr_off,
      rush_att   = season_rush_attempts__yr_off,
      rush_tds   = season_rush_touchdown__yr_off,
      recs       = season_receptions__yr_off,
      rec_yards  = season_receiving_yards__yr_off,
      rec_tds    = season_receiving_touchdown__yr_off,
      log_value  = log(value__cmb)
    ) %>%
    drop_na(log_value, age, years_exp)

  if (nrow(dat_pos) < 40) {
    message("Not enough observations for ", pos_label, " after cleaning; skipping.")
    return(NULL)
  }

  dat_pos <- dat_pos %>%
    mutate(
      age_z        = standardize(age),
      exp_z        = standardize(years_exp),
      pass_yards_z = standardize(pass_yards),
      pass_tds_z   = standardize(pass_tds),
      ints_z       = standardize(ints),
      pass_att_z   = standardize(pass_att),
      rush_yards_z = standardize(rush_yards),
      rush_att_z   = standardize(rush_att),
      rush_tds_z   = standardize(rush_tds),
      recs_z       = standardize(recs),
      rec_yards_z  = standardize(rec_yards),
      rec_tds_z    = standardize(rec_tds)
    )

  # position‑specific formulas
  formula_pos <- switch(
    pos_label,
    "QB" = log_value ~ age_z + exp_z +
      pass_yards_z + pass_tds_z + ints_z + pass_att_z,
    "RB" = log_value ~ age_z + exp_z +
      rush_yards_z + rush_att_z + rush_tds_z + recs_z + rec_yards_z,
    "WR_TE" = log_value ~ age_z + exp_z +
      rec_yards_z + recs_z + rec_tds_z,
    # default
    log_value ~ age_z + exp_z
  )

  splits <- split_three_way(dat_pos, train_prop = 0.6, test_prop = 0.2, seed = 340)

  fit <- lm(formula_pos, data = splits$train)

  # metrics
  test_pred <- predict(fit, newdata = splits$test)
  val_pred  <- predict(fit, newdata = splits$val)

  test_metrics <- regression_metrics(splits$test$log_value, test_pred) %>%
    mutate(set = "test", pos_group = pos_label)
  val_metrics  <- regression_metrics(splits$val$log_value,  val_pred) %>%
    mutate(set = "validation", pos_group = pos_label)

  list(
    fit     = fit,
    splits  = splits,
    metrics = bind_rows(test_metrics, val_metrics)
  )
}

pos_groups <- c("QB", "RB", "WR_TE")
off_results <- lapply(pos_groups, function(pg) fit_position_model(offense_pos, pg))
names(off_results) <- pos_groups

off_metrics <- off_results %>%
  purrr::discard(is.null) %>%
  purrr::map("metrics") %>%
  bind_rows()

off_metrics
```

### Interpretation

- **RMSE** is the typical prediction error in log(contract value).  
  Lower bars indicate more accurate salary predictions for that position group.
- **R²** measures how much of the variance in log(contract value) the model
  explains. Higher R² means performance stats do a better job of separating
  higher‑paid from lower‑paid players at that position.
- WR/TE, has noticeably weaker R² — that suggests contract decisions for that group depend    more on factors not captured in basic box‑score stats.

# 5. Defensive Regression Model

```{r}
defense_model <- defense_clean %>%
  mutate(
    age_z      = standardize(age),
    exp_z      = standardize(years_exp),
    tackles_z  = standardize(total_tackles),
    sacks_z    = standardize(sacks),
    qb_hits_z  = standardize(qb_hits),
    ints_z     = standardize(ints)
  )

def_splits <- split_three_way(defense_model, train_prop = 0.6, test_prop = 0.2, seed = 340)

def_fit <- lm(
  log_value ~ age_z + exp_z + tackles_z + sacks_z + qb_hits_z + ints_z,
  data = def_splits$train
)

def_test_pred <- predict(def_fit, newdata = def_splits$test)
def_val_pred  <- predict(def_fit, newdata = def_splits$val)

def_metrics <- bind_rows(
  regression_metrics(def_splits$test$log_value, def_test_pred) %>%
    mutate(set = "test", pos_group = "DEFENSE"),
  regression_metrics(def_splits$val$log_value,  def_val_pred) %>%
    mutate(set = "validation", pos_group = "DEFENSE")
)

def_metrics
```

# 6. Combined Metrics Table and Plots

```{r}
all_metrics <- bind_rows(
  off_metrics %>% mutate(model_type = "Offense"),
  def_metrics %>% mutate(model_type = "Defense")
) %>%
  mutate(
    pos_group = factor(pos_group, levels = c("QB", "RB", "WR_TE", "DEFENSE")),
    set = factor(set, levels = c("test", "validation"))
  )

kable(
  all_metrics %>% arrange(model_type, pos_group, set),
  digits = 3,
  caption = "RMSE and R² by Position Group and Data Split"
)
```

```{r}
ggplot(all_metrics, aes(x = pos_group, y = RMSE, fill = set)) +
  geom_col(position = "dodge") +
  labs(
    title = "RMSE by Position Group and Data Split",
    x = "Position Group",
    y = "RMSE (log contract value)"
  ) +
  theme_minimal()
```

```{r}
ggplot(all_metrics, aes(x = pos_group, y = R2, fill = set)) +
  geom_col(position = "dodge") +
  labs(
    title = "R² by Position Group and Data Split",
    x = "Position Group",
    y = "R²"
  ) +
  theme_minimal()
```

### How to Read These Plots

- If **QB** has the highest R² and relatively low RMSE, that supports the idea
  that quarterback salaries track recent performance pretty closely.
- If **RB** shows moderate R² with somewhat higher RMSE, it suggests teams
  consider rushing and receiving production but also weigh other factors like
  age, workload, and injury risk.
- If **WR/TE** has low or even negative R², that indicates our simple box‑score
  stats are not very predictive of contract size for pass‑catchers — an
  interesting finding you can discuss as a limitation and an avenue for future
  work.
- Defensive results summarize how much basic counting stats for defenders
  explain variation in defensive contract values.

# 7. Coefficient Tables and Visualizations

## 7.1 Offensive Coefficients

```{r}
get_coef_df <- function(res_obj, label) {
  if (is.null(res_obj)) return(NULL)
  broom::tidy(res_obj$fit) %>%
    filter(term != "(Intercept)") %>%
    mutate(pos_group = label)
}

off_coefs <- bind_rows(
  get_coef_df(off_results$QB, "QB"),
  get_coef_df(off_results$RB, "RB"),
  get_coef_df(off_results$WR_TE, "WR_TE")
)

kable(
  off_coefs %>%
    select(pos_group, term, estimate, std.error, statistic, p.value) %>%
    arrange(pos_group, desc(abs(estimate))),
  digits = 3,
  caption = "Offensive Regression Coefficients (Standardized Predictors)"
)
```

```{r}
ggplot(off_coefs, aes(x = term, y = estimate)) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_point() +
  geom_errorbar(
    aes(ymin = estimate - 1.96 * std.error,
        ymax = estimate + 1.96 * std.error),
    width = 0.1
  ) +
  coord_flip() +
  facet_wrap(~ pos_group, scales = "free_y") +
  labs(
    title = "Standardized Coefficients by Offensive Position Group",
    x = "Predictor (z‑score)",
    y = "Effect on log(contract value)"
  ) +
  theme_minimal()
```

### Interpretation

Because all predictors are standardized (z‑scores), each coefficient tells you:

> “Holding other variables fixed, a one‑standard‑deviation increase in this
> stat is associated with a `coef × 100%` change in contract value.”


- For QBs, a large positive coefficient on **passing touchdowns** means higher
  TD totals are strongly associated with larger contracts.
- For RBs, both **rushing yards** and **receiving yards**
  matter, reflecting the league’s preference for versatile backs.
- For WR/TE, small or noisy coefficients reinforce that contracts for
  pass‑catchers depend on more than just basic volume stats.

## 7.2 Defensive Coefficients

```{r}
def_coefs <- broom::tidy(def_fit) %>%
  filter(term != "(Intercept)") %>%
  mutate(pos_group = "DEFENSE")

kable(
  def_coefs %>%
    select(term, estimate, std.error, statistic, p.value),
  digits = 3,
  caption = "Defensive Regression Coefficients (Standardized Predictors)"
)
```

```{r}
ggplot(def_coefs, aes(x = term, y = estimate)) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_point() +
  geom_errorbar(
    aes(ymin = estimate - 1.96 * std.error,
        ymax = estimate + 1.96 * std.error),
    width = 0.1
  ) +
  coord_flip() +
  labs(
    title = "Standardized Coefficients for Defensive Model",
    x = "Predictor (z‑score)",
    y = "Effect on log(contract value)"
  ) +
  theme_minimal()
```

